Towards planning of infantile cardiac surgery based on simulation

### 2012-02-29 Doc/Miccai/shell-model.tex
changes to simulation and shell model sections
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@349 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

In addition, the computation of this model must be fast enough to allows for interactive planning of the surgery. 
The options on the choice of model are limited: 
The model based on the deformation of the centerline, that can be very quickly computed, (like ??\tomas{Add refs.!} \christian{Do you have some refs that uses model based on centerline? }  ) must be discarded because they do not support cut of the wall along the vessel axis. 
Mass-springs approaches that are often used for cloth simulation (\christian{do we consider this option ? if yes add ref ??}), allow for fast computation but do not provide sufficient accuracy as they do not necessary converge to a unique solution when changing the mesh.
In this work, we then propose to rely on Finite Element formulation of thin-shell theory. 
These model are numerically more expensive, but provide accurate models of thin deformable structure.% that converge to unique solutions.
Moreover, we show that we can still obtain fast computation with a reasonable number of elements.

\subsubsection{Thin shell elements: } 
The formulation of the thin shell elements is based on the mechanics of continuum. 
\tomas{image: single shell}
For thorough description we refer the reader to the textbook about Finite Element methods for example \cite{Reddy1993}.
These elements are commonly used in physics and mechanics for deformation modelling of thin structures because they are numerically more stable then volumetric FEM methods in this situation. 
The shell element is a 2-dimensional element (in our case, triangle element) whose thickness is a parameter of the simulation.
Often the thickness is considered constant over the surface of the element.
Shell combines forces from in-plane deformations, like stretching or shearing, with bending forces. 
Such internal forces $\myvec{f}$ are a function of current displacement measured as difference of current shape
$\myvec{x}$ from the rest shape $\myvec{x_0}$.
To accelerate the computation, we use a corotational formulation, that is based on the definition of a local frame for each element (that can be built using the geometry of the triangle, \christian{see figure "image: single shell" ?? }).

In this local frame, the computation of the internal forces is linear:
\begin{equation}
  \myvec{f} = \mathbf{K_e}(\myvec{x} - \myvec{x_0})
 \label{linear_static_model} 
\end{equation}
where $\mathbf{K_e}$ is a stiffness matrix that is computed thanks to a weak formulation (integral formulation) of Kirshohoff-Loves equations.
This approach is very similar to the model presented in  \cite{Comas2010b,Comas2010c}  even if the interpolation functions we use are  different to allow for better convergence\footnote{Our shell model will be published separately}. 
We emphasis that the method described in the paper is independent of the shell model used (other non-corotational shell models can be used).
%
When the structure is assembled, the contributions of each element are computed in the global frame using a different rotation matrix for each element.
\begin{equation}
\mathbb{K} = \mathcal{A}_{e=0}^N (\mathbf{R_e}^T \mathbf{K_e} \mathbf{R_e}) \hspace{10mm}
\mathbb{F} = \mathcal{A}_{e=0}^N ( \mathbf{R_e}^T \mathbf{K_e}(\myvec{x} - \myvec{x_0}))
\end{equation}
Where $\mathcal{A}$ is the assembling operator over the $N$ elements of the mesh, $\mathbb{K}$ is the global stiffness and $\mathbb{F}$ the internal forces of the entire shell mesh structure. 
Using this approach, $\mathbb{K}$ is changing at each step of the simulation but the model is able to catch geometric non-linear deformations.
In practice, the computation of the node motion $\mathbb{U} = \mathbb{K}^{-1} \mathbb{F}$ is performed using an optimized conjugate gradient algorithm.

@@

Figure \ref{fig-deformations} shows different deformations that are likely to
occur during the surgery (bending and twisting). The figure compares our
results with the results of Li et al. \cite{Li2009}. As you can see the
results are close to the real deformations.

@@

\begin{figure}[tbh]
  \centering
  \begin{tabular}{cccc}
    \includegraphics[width=0.24\columnwidth]{img/twist-06-cg.png}
    &
    \includegraphics[width=0.24\columnwidth]{img/twist-08-cg.png}
    &
    \includegraphics[width=0.24\columnwidth]{img/twist-16-cg.png}
    &
    \includegraphics[width=0.24\columnwidth]{img/twist-31-cg.png}
    \\
    \includegraphics[width=0.24\columnwidth]{img/twist-06w-cg.png}
    &
    \includegraphics[width=0.24\columnwidth]{img/twist-08w-cg.png}
    &
    \includegraphics[width=0.24\columnwidth]{img/twist-16w-cg.png}
    &
    \includegraphics[width=0.24\columnwidth]{img/twist-31w-cg.png}
  \end{tabular}
  \caption{Convergence on a twisted cylinder with 6, 8, 16 and 31 nodes
  around the circumference (120, 208, 832 and 3038 elements).
  \tomas{improve the wireframe images if there's time} } 
  \label{fig-convergence}
\end{figure}

\christian{ I think this should be put in the result section:
Since we use the method as a
geometrical tool we have also certain independence on physical parameters.
For example the variation of Young's modulus in the interval from $10^3$ to
$10^7$ resulted to only $0.1\%$ change of the deformation.
}

% vim: et sw=2 tw=75 fdm=marker fdc=2 spell

>>>>

In addition, the computation of this model must be fast enough to allows for interactive planning of the surgery. 
The options on the choice of model are limited: 
The model based on the deformation of the centerline, that can be very quickly computed, (like ??\tomas{Add refs.!} \christian{Do you have some refs that uses model based on centerline? }\tomas{No, but I will find something}  ) must be discarded because they do not support cut of the wall along the vessel axis. 
Mass-springs approaches that are often used for cloth simulation \cite{Volino2005b}, allow for fast computation but do not provide sufficient accuracy as they do not necessary converge to a unique solution when changing the mesh.
In this work, we then propose to rely on Finite Element formulation of thin-shell theory. 
These model are numerically more expensive, but provide accurate models of thin deformable structure.% that converge to unique solutions.
Moreover, we show that we can still obtain fast computation with a reasonable number of elements.

\subsubsection{Thin shell elements: } 
The formulation of the thin shell elements is based on the mechanics of continuum. 
For thorough description we refer the reader to the textbook about Finite Element methods for example \cite{Reddy1993}.
These elements are commonly used in physics and mechanics for deformation modelling of thin structures because they are numerically more stable then volumetric FEM methods in this situation. 
The shell element is a 2-dimensional element (in our case, triangular element) whose thickness is a parameter of the simulation (often considered constant over the surface of the element).
Shell element combines forces from in-plane deformations, like stretching or shearing, with bending forces. 
Such internal forces $\myvec{f}$ are a function of current displacement measured as difference of current shape
$\myvec{x}$ from the rest shape $\myvec{x_0}$.
To accelerate the computation, we use a corotational formulation, that is based on the definition of a local frame for each element (that can be built using the geometry of the triangle, \christian{see figure "image: single shell" ?? }).

In this local frame, the computation of the internal forces is linear:
\begin{equation}
  \myvec{f} = \mathbf{K_e}(\myvec{x} - \myvec{x_0})
 \label{linear_static_model} 
\end{equation}
where $\mymat{K_e}$ is a stiffness matrix that is computed thanks to a weak formulation (integral formulation) of Kirshohoff-Loves equations.
This approach is very similar to the model presented in  \cite{Comas2010b,Comas2010c}  even if the interpolation functions we use are  different to allow for better convergence\footnote{Our shell model will be published separately}. 
We emphasize that the method described in the paper is independent of the shell model used (other non-corotational shell models can be used as well).
%
When the structure is assembled, the contributions of each element are computed in the global frame using a different rotation matrix for each element.
\begin{equation}
\mathbb{K} = \mathcal{A}_{e=0}^N (\mathbf{R_e}^T \mathbf{K_e} \mathbf{R_e}) \hspace{10mm}
\mathbb{F} = \mathcal{A}_{e=0}^N ( \mathbf{R_e}^T \mathbf{K_e}(\myvec{x} - \myvec{x_0}))
\end{equation}
Where $\mathcal{A}$ is the assembling operator over the $N$ elements of the mesh, $\mathbb{K}$ is the global stiffness and $\mathbb{F}$ the internal forces of the entire shell mesh structure. 
Using this approach, $\mathbb{K}$ is changing at each step of the simulation but the model is able to catch geometric non-linear deformations.
In practice, the computation of the node motion $\mathbb{U} = \mathbb{K}^{-1} \mathbb{F}$ is performed using an optimized conjugate gradient algorithm.

@@

Figure \ref{fig-deformations} shows bending and twisting deformations.
These are likely to occur during the surgery.
The figure compares our
results with the results of Li et al. \cite{Li2009}. As you can see the
results are close to the real deformations.

@@

\begin{figure}[tbh]
  \centering
  \begin{tabular}{cccc}
    \includegraphics[width=0.24\columnwidth]{img/twist-06-cg.png}
    &
    \includegraphics[width=0.24\columnwidth]{img/twist-08-cg.png}
    &
    \includegraphics[width=0.24\columnwidth]{img/twist-16-cg.png}
    &
    \includegraphics[width=0.24\columnwidth]{img/twist-31-cg.png}
    \\
    \includegraphics[width=0.24\columnwidth]{img/twist-06w-cg.png}
    &
    \includegraphics[width=0.24\columnwidth]{img/twist-08w-cg.png}
    &
    \includegraphics[width=0.24\columnwidth]{img/twist-16w-cg.png}
    &
    \includegraphics[width=0.24\columnwidth]{img/twist-31w-cg.png}
  \end{tabular}
  \caption{Convergence on a twisted cylinder with 6, 8, 16 and 31 nodes
  around the circumference (120, 208, 832 and 3038 elements).
  Using mesh with 3038 triangles for visualisation.
  \tomas{improve the wireframe images if there's time} } 
  \label{fig-convergence}
\end{figure}

\christian{ I think this should be put in the result section:
Since we use the method as a
geometrical tool we have also certain independence on physical parameters.
For example the variation of Young's modulus in the interval from $10^3$ to
$10^7$ resulted in only $0.1\%$ change of the deformation.
}

% vim: et sw=2 tw=0 spell

### 2012-02-29 Doc/Miccai/simulation.tex
changes to simulation and shell model sections
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@349 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

There are several types of low-level task the surgeon usually employs while
performing the surgery.

One common procedure is reconnection and joining of blood vessels. Another
is fixing the stenosis of aorta where the three most common types of fixing
the problem are:

\begin{description}

@@

From our point of view the all these procedures share the same similarity:
parts of the circulatory system are sutured together or to a patch. These
parts are not too far away from each other. On this we have based our
general approach in the simulation. The simple way how to perform the
suturing would be to connect the respective edges with springs. This
approach however adds some unwanted energy to the system. Our method is
based solely on the relaxation of the mesh of shell elements.

We start with a mesh representing the situation after all the cuts have
been performed but before suturing takes place. In the first step we select
on the mesh edges that we want to suture together and join the
corresponding nodes together. The result is topologicaly different mesh
that has same number of elements. The Figure. \ref{fig-JoiningVessels} explains
this operation. Then we start the simulation on the modified mesh while
using the original mesh to define rest shapes of the shell elements. After
the mesh relaxes and the simulation reaches energetical equilibrium the
simulation terminates.

@@

\caption{Scheme showing the method for joining two vessels.}
\label{fig-JoiningVessels}
\end{figure}

If the nodes that are to be sutured together are not in a close vicinity we
may experience unwanted quick increase of internal forces at the beginning
of the simulation. To allevieate this problem we have implemented the
techinuqe of linear interpolation of the rest shape mesh. At the start of
the simulation we use different rest shape mesh. It is topologicaly
equivalent to the original mesh but nodes have positions equivalent to the
simulated mesh (that way the initial deformations are not so big). During
the simulation we then perform several steps of linear interpolation
between these two rest shape meshes. At the end of interpolation the
original rest shape is used. After the mesh relaxes the simulation again
terminates.

% vim: et sw=2 tw=75 fdm=marker fdc=2 spell

>>>>

There are several types of low-level task the surgeon usually employs while
performing the surgery.
The most common:

\begin{description}
  \item Reconnection and joining of blood vessels to the arteries.

@@

With our method we are able to simulate all these tasks.
We have based our general approach on the common task involved in all these operations which is suturing of two or more edges together.
Since we are interested in predicting the result rather than creating a training simulator for suturing (trained surgeon knows well how to perform the sutures) the simplest way of performing the suture would be to connect the respective edges with springs. 
Springs however adds some unwanted energy to the system. 
Our method is based solely on the relaxation of the mesh of shell elements.
%% christian: I like the "relaxation" term !!!

We start with a mesh representing the situation after all the cuts have been performed but before suturing takes place. 
In the first step we perform the required topological changes to the mesh. We keep the number of elements in new mesh the same as in the original mesh so that we can define 1:1 mapping between elements of these meshes.
The Figure \ref{fig-JoiningVessels} explains this operation when joining vessels. 
Then we start the simulation on the modified mesh while using the original mesh to define rest shapes of all shell elements.

@@

\caption{Scheme showing the method for joining two vessels. \christian{Do we need to extend the scheme for every type of aortoplasty (end to end / Patch / Subclavian): it would illustrate that the relaxation method works for every case...}
  \tomas{No we don't. This there will be real examples in results (if I remember correctly).} }
\label{fig-JoiningVessels}
\end{figure}

In case of non-trivial distance between the nodes that are sutured together the involved elements will undergo large deformation at the beginning of the simulation.
This results in quick increase of internal forces which can cause unwanted effects.
To allevieate the problem we have implemented the technique that allows for moderate changes in the rest shape between simulation steps.
Between the start and the end of the simulation we use a linear interpolation of the rest shape for the shells that lie along the sutured edges. 
\christian{I think that it is unclear that we do the relaxation on some shells that are on the side edges of the connecting line}
\tomas{I don't understand what do you mean by that. We will have to discuss that.}
The interpolation starts with the rest shape equal to the deformed elements and we progressively recontract the rest shape of the relaxed shells to their initial shapes. 
The energetical equilibrium found by the simulation gradually forces the side edges to connect.
After the mesh relaxes the simulation again terminates.

% vim: et sw=2 tw=0 spell
